# One way ANOVA

## ANOVA Table

Let us consider an example where we have results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions. The dataset is named as `PlantGrowth` and it can be found in the library `car`.

```{r, message=FALSE}
# Loading the dataset
library(car)
library(DescTools)
library(tidyverse)
library(lmtest)
data <- PlantGrowth
```

Let us print the anova table.

```{r}
anova(lm(weight ~ group, data = data))
```

The null hypothesis that all treatments are the same is rejected at a significance level of 5\%.

::: {.callout-note appearance="simple"}

**Manual calculation:** Let us perform the calculations by the `anova()` manually for better understanding.

```{r}
# Finding the treatment means
treatment_means <- data %>% 
  group_by(group) %>% 
  summarise(mean_wt = mean(weight))

# Replicates
n <- 10

# Treatments
k <- 3

# Total experiments
N <- k*n

# grand mean
grand_mean <- mean(data$weight)

# Error sum of squares
data_with_treatment_mean <- merge(data, treatment_means)
SSE <- sum((data_with_treatment_mean$weight - data_with_treatment_mean$mean_wt)**2)
print(paste0("SSE = ", SSE))

# Treatment sum of squares
SSTr <- n*sum((treatment_means$mean_wt - grand_mean)**2)
print(paste0("SSTr = ", SSTr))

# Mean sum of squared error
MSE <- SSE / (N - k)
print(paste0("MSE = ", MSE))

# Mean sum of square treatment
MSTr <- SSTr / (k - 1)
print(paste0("MSTr = ", MSTr))

# F-statistic
F <- MSTr / MSE
print(paste0("F-statistic = ", F))

# p-value
pf(F, k-1, N-k, lower.tail = FALSE)
```
:::

## Confidence interval

Let us print the Bonferroni's confidence intervals:

```{r}
aov_object <- aov(lm(weight ~ group, data = data))
PostHocTest(aov_object, method = "bonferroni")
```

Based on Bonferroni's method, we observe that treatment 1 and treatment 2 are significantly different, but there is no difference between other pairs.

::: {.callout-note appearance="simple"}

**Manual calculation:** Let us perform the calculations by the `PostHocTest(method = "bonferroni")` function manually for better understanding.

```{r}
# Estimates of the pairwise difference of treatment means

print(paste0("Estimated difference between treatment 2 and treatment 1 = ", treatment_means$mean_wt[2] - treatment_means$mean_wt[1]))
print(paste0("Estimated difference between treatment 3 and treatment 1 = ", treatment_means$mean_wt[3] - treatment_means$mean_wt[1]))
print(paste0("Estimated difference between treatment 3 and treatment 2 = ", treatment_means$mean_wt[3] - treatment_means$mean_wt[2]))
```

```{r}
# Calculating the Bonferroni's confidence interval

# Significance level
alpha = 0.05

# Number of comparisons
k_dash = 3

# Upper bound of the confidence interval for the estimated difference between treatment 2 and treatment 1 

(treatment_means$mean_wt[2] - treatment_means$mean_wt[1]) + (sqrt(MSE))*sqrt(2/n)*qt(1 - alpha/(2*k_dash), N - k)
```

```{r}
# Lower bound of the confidence interval for the estimated difference between treatment 2 and treatment 1 

(treatment_means$mean_wt[2] - treatment_means$mean_wt[1]) - (sqrt(MSE))*sqrt(2/n)*qt(1 - alpha/(2*k_dash), N - k)
```

Similarly, we can calculate the Bonferroni's confidence interval of other estimates.
:::


Bonferroni's confidence intervals are overly conservative. Let us find the confidence intervals basead on Tukey's method:

```{r}
PostHocTest(aov_object, method = "hsd")
```

::: {.callout-note appearance="simple"}

**Manual calculation:** Let us perform the calculations by the `PostHocTest(method = "hsd")` function manually for better understanding.

The calculation of the estimates of the pairwise difference of treatment means is the same as that for Bonferroni's method.

```{r}
# Calculating the Tukey's confidence interval

# Significance level
alpha = 0.05

# Number of comparisons
k_dash = 3

# Upper bound of the confidence interval for the estimated difference between treatment 2 and treatment 1 

(treatment_means$mean_wt[2] - treatment_means$mean_wt[1]) + (sqrt(MSE))*sqrt(2/n)*(1/sqrt(2))*qtukey(1 - alpha, nmeans = 3, df = N - k)
```

```{r}
# Lower bound of the confidence interval for the estimated difference between treatment 2 and treatment 1 

(treatment_means$mean_wt[2] - treatment_means$mean_wt[1]) - (sqrt(MSE))*sqrt(2/n)*(1/sqrt(2))*qtukey(1 - alpha, nmeans = 3, N - k)
```

Similarly, we can calculate the Tukey's confidence interval of other estimates.
:::

## Model assumptions check

The conclusions of the statistical tests and the confidence intervals are based on the assumption that random error is normally distributed with mean 0 and constant variance, i.e., $\epsilon_{ij} \sim N(0, \sigma^2)$. 

The diagnostic plots and statistical tests for checking model assumptions can be found [here](https://nustat.github.io/STAT350-class-notes/Diagnostics_Remedial_Measures.html#diagnostic-plots-statistical-tests).

## Contrasts

### Class comparison

This is an example to use orthogonal contrasts to answer questions of interest. The data `rice_seed_data.csv` consists of the shoot dry weight (in mg) of an experiment to determine the effect of seed treatment by acids on the early growth of rice seeds.

The investigator had several specific questions in mind from the beginning: 
 -Do acid treatments affect seedling growth?
- Is the effect of organic acids different from that of inorganic acids?
- Is there a difference in the effects of the two different organic acids?

We will create orthogonal contrasts, and perform an ANOVA analysis to answer these questions.


```{r}
rice_data <- read.csv('rice_seed_data.csv')
anova(lm(growth~treatment, data = rice_data))

c1 <- c(3, -1, -1, -1)
c2 <- c(0, -2, 1, 1)
c3 <- c(0, 0, 1, -1)

# Contrast matrix
mat.contrast <- cbind(c1, c2, c3)
colnames(mat.contrast) <- paste0("c",1:3)

# Converting treatment to a factor variable
rice_data$treatment <- as.factor(rice_data$treatment)
contrasts(rice_data$treatment) <- mat.contrast

# ANOVA table
model <- aov(growth ~ treatment, data = rice_data,
                   contrasts = list(treatment = mat.contrast))

# Splitting the Treatment sum of squares into independent 
# orthogonal contrast components
summary.aov(model,split = list(treatment = list("Control vs acid"=1,  
                              "Inorganic vs organic" = 2, 'Between organic'=3)))

```

We conclude that there is no statistically significant difference between the effect of organic acids. However, there is a statistically significant difference between the effect of inorganic and organic acids, and an even higher statistically significant difference between the weight of seeds in the control group and the weight seeds treated with acid.

### Model assumptions check

```{r}
par(mfrow = c(2,2))
plot(model)
```

Based on the plot of the residuals against fitted values, the average residual is close to zero. Thus, the assumption of the error term having a zero mean is satisfied.

```{r}
bptest(model)
```

Based on the Breusch Pagan test, the homoscedasticity assumption is satisfied.

```{r}
shapiro.test(model$residuals)
```

Based on Shapiro’s test, the residuals are normally distributed.

Thus, all the model assumptions are satisfied.

### Trend comparison

This is an example to identify if there is a linear, quadratic, or higher order relationship between a continuous treatment and the response.

Consider the data `laser_power.csv`, which consists of laser power, and the corresponding strength of the material obtained using a machining process involving the laser. The question to be answered is if there is a linear or quadratic relationship between the laser power and strength of the material.

```{r}
data <- read.csv('laser_power.csv')
data$power <- as.factor(data$power)

# Linear contrast
c1 <- c(-1, 0, 1)

# Quadratic contrast
c2 <- c(1, -2, 1)

mat.contrast <- cbind(c1, c2)
model <- aov(strength ~ power, data = data,
             contrasts = list(power = mat.contrast))
summary.aov(model, split = list(power = list("linear" = 1,
                                                 "quadratic" = 2)))
```

We conclude that there is a linear relationship between laser power and strength. This can be seen visually as well as below.

```{r}
data$power <- as.integer(substr(data$power,5,8))
plot(data$power, data$strength)
abline(lm(strength~power, data = data))
```


