# One way ANOVA

## ANOVA Table

Let us consider an example where we have results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions. The dataset is named as `PlantGrowth` and it can be found in the library `car`.

```{r, message=FALSE}
# Loading the dataset
library(car)
library(DescTools)
data <- PlantGrowth
```

Let us print the anova table.

```{r}
anova(lm(weight ~ group, data = data))
```

The null hypothesis that all treatments are the same is rejected at a significance level of 5\%.

## Confidence interval

Let us print the Bonferroni's confidence intervals:

```{r}
aov_object <- aov(lm(weight ~ group, data = data))
PostHocTest(aov_object, method = "bonferroni")
```

Based on Bonferroni's method, we observe that treatment 1 and treatment 2 are significantly different, but there is no difference between other pairs.

Bonferroni's confidence intervals are overly conservative. Let us find the confidence intervals basead on Tukey's method:

```{r}
PostHocTest(aov_object, method = "hsd")
```

## Model assumptions check

The conclusions of the statistical tests and the confidence intervals are based on the assumption that random error is normally distributed with mean 0 and constant variance, i.e., $\epsilon_{ij} \sim N(0, \sigma^2)$. 

The diagnostic plots and statistical tests for checking model assumptions can be found [here](https://nustat.github.io/STAT350-class-notes/Diagnostics_Remedial_Measures.html#diagnostic-plots-statistical-tests).

## Contrasts

This is an example to use orthogonal contrasts to answer questions of interest. The data `rice_seed_data.csv` consists of the shoot dry weight (in mg) of an experiment to determine the effect of seed treatment by acids on the early growth of rice seeds.

The investigator had several specific questions in mind from the beginning: 
 -Do acid treatments affect seedling growth?
- Is the effect of organic acids different from that of inorganic acids?
- Is there a difference in the effects of the two different organic acids?

We will create orthogonal contrasts, and perform an ANOVA analysis to answer these questions.


```{r}
rice_data <- read.csv('rice_seed_data.csv')
anova(lm(growth~treatment, data = rice_data))

c1 <- c(3, -1, -1, -1)
c2 <- c(0, -2, 1, 1)
c3 <- c(0, 0, 1, -1)

# Contrast matrix
mat.contrast <- cbind(c1, c2, c3)
colnames(mat.contrast) <- paste0("c",1:3)

# Converting treatment to a factor variable
rice_data$treatment <- as.factor(rice_data$treatment)
contrasts(rice_data$treatment) <- mat.contrast

# ANOVA table
model <- aov(growth ~ treatment, data = rice_data,
                   contrasts = list(treatment = mat.contrast))

# Splitting the Treatment sum of squares into independent 
# orthogonal contrast components
summary.aov(model,split = list(treatment = list("Control vs acid"=1,  
                              "Inorganic vs organic" = 2, 'Between organic'=3)))

```

We conclude that there is no statistically significant difference between the effect of organic acids. However, there is a statistically significant difference between the effect of inorganic and organic acids, and an even higher statistically significant difference between the weight of seeds in the control group and the weight seeds treated with acid.


