# One way ANOVA

## ANOVA Table

Let us consider an example where we have results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions. The dataset is named as `PlantGrowth` and it can be found in the library `car`.

```{r, message=FALSE}
# Loading the dataset
library(car)
library(DescTools)
data <- PlantGrowth
```

Let us print the anova table.

```{r}
anova(lm(weight ~ group, data = data))
```

The null hypothesis that all treatments are the same is rejected at a significance level of 5\%.

Let us print the Bonferroni's confidence intervals:

```{r}
aov_object <- aov(lm(weight ~ group, data = data))
PostHocTest(aov_object, method = "bonferroni")
```

Based on Bonferroni's method, we observe that treatment 1 and treatment 2 are significantly different, but there is no difference between other pairs.

Bonferroni's confidence intervals are overly conservative. Let us find the confidence intervals basead on Tukey's method:

```{r}
PostHocTest(aov_object, method = "hsd")
```





