# Computer experiment designs

## Maximin design

Let us generate a 4-run maximin design in $[0, 1]^2$ for $p=2$.

We will use the `maximin()` function of the `maximin` library

```{r, message = FALSE}
library(maximin)
```

For generating a maximin design, we need to specify an initial design, and the function will add points to the design to develop a design that is as close to a maximin design as possible.

Let us add the origin as a point in the initial design.

```{r}
D_initial <- matrix(c(0,0),1,2)
```

```{r}
set.seed(3)
D <- maximin(n = 3, p=2, Xorig = D_initial)$Xf
```

```{r}
plot(D, pch = 16)
```

As expected, the design is the four corners of the feasible design space as the minimum distance between any two design points is the largest in this case.

## Maximin LHD

Let us generate a 4-run maximin LHD in $[0, 1]^2$ for $p=2$.

We will use the `maximinLHS()` function of the `lhs` library

```{r, message = FALSE}
library(lhs)
```

```{r}
set.seed(1)
D <- maximinLHS(4, 2)
plot(D, xlim = c(0, 1), ylim = c(0, 1), pch = 16)
```

Note that the design has 4 distinct projections for each of the two factors.



