# Diagnostics and Remedial measures

Let us consider the `Auto` dataset from the `ISLR` library. The library is related to the book *[Introduction to Statistical Learning](https://hastie.su.domains/ISLR2/ISLRv2_website.pdf)*

## Loading libraries

```{r}
library(ISLR)
library(ggplot2)
library(lmtest)
```


## Reading data

```{r}
auto_data <- Auto
```

## Developing model

Suppose we wish to predict `mpg` based on `horsepower`.

```{r}
linear_model <- lm(mpg~horsepower, data = auto_data)
```

## Model fit

Let us visualize the model fit to the data.

```{r}
ggplot(data = auto_data, aes(x = horsepower, y = mpg))+
  geom_point()+
  geom_smooth(method = "lm")
```

The plot indicates that a quadratic relationship between `mpg` and `horsepower` may be more appropriate. 

## Diagnostic plots

Let us make diagnostic plots that help us check the model assumptions.

```{r}
par(mfrow = c(2,2))
plot(model)
```

### Linearity assumption

#### Visual check

We can check the linearity assumption visually by analyzing the plot of residuals against fitted values. The plot indicates that a quadratic relationship between the response and the predictor may be more appropriate than a linear one.

#### Statistical test

We'll use the *F* test for lack of fit to check the linearity assumption

```{r}
#Full model
full_model <- lm(mpg~as.factor(horsepower), data = auto_data)
```

```{r}
# F test for lack of fit
anova(linear_model, full_model)
```

As expected based on the visual check, the linear model fails the linearity test with a very low $p$-value of the order of $10^{-14}$. Thus, we conclude that relationship is non-linear.

### Constant variance

#### Visual check

The plot of the residuals against fitted values indicates that the error variance is increaing with increasing values of predicted `mpg` or decreasing `horsepower`. Thus, there seems to be heteroscedasticity.

#### Statistical test

Let us conduct the Breusch-Pagan test for homoscedasticity. This is a large sample test, which assumes that the error terms are independent and normally distruted, and that the variance of of the error term $\epsilon_i$, denoted by $\sigma_i^2$, is related to the level of the predictor $X$ in the following way:

$\log(\sigma_i^2) = \gamma_0 + \gamma_1X_i$.

This implies that $\sigma_i^2$ either increases or decreases with the level of $X$. Constant error variance corresponds to $\gamma_1 = 0$. The hypotheses for the test are:

$H_0: \gamma_1 = 0$
$H_a: \gamma_1 \ne 0$

```{r}
bptest(linear_model)
```

As expected based on the visual check, the model fails the homoscedasticity with a low $p$-value of 0.3\%.



