# Multiple factors

## Randomized complete block design (RCBD)

Let us consider an example, where the shear strength of steel plate girders needs to be modeled as a function of the four methods *(treatment)* and nine girders *(blocks)*.

```{r}
library(ACSWR)
library(reshape2)
library(DescTools)
library(lmtest)

# Visualizing treatment effects
data("girder")
boxplot(girder[,2:5])

# Melting data to make it suitable to fit a linear regression model
# using the 'lm' function
data_melt <- melt(girder, variable.name = "treatment", id = 'Girder')
anova(lm(value~Girder+treatment, data = data_melt))
```

The null hypothesis that there is no effect of the method on the sheer strength of the girder is rejected.

Let us compare the methods pairwise.

```{r}
aov_object <- aov(value~Girder+treatment, data = data_melt)

# Tukey's method
PostHocTest(aov_object, method = "hsd")$treatment
```

All methods are different from each other according to Tukey's method *(at 5\% significance level)*.


```{r}
# Bonferroni's method
PostHocTest(aov_object, method = "bonferroni")$treatment
```

All methods except Aarau and Cardiff are different from each other according to Bonferroni's method *(at 5\% significance level)*. 

Let us verify if the model assumptions are satisfied.

```{r}
par(mfrow = c(2,2))
model <- lm(value~Girder+treatment, data = data_melt)
plot(model)
```


```{r}
shapiro.test(model$residuals)
```

The errors are normally distribued.

```{r}
bptest(model)
```

The error variance assumption is also satisfied at a 1\% significance level. There is no strong deviation from the assumption.


## Two-way layout: Fixed effects

A manufacturer was interested in finding differences in torque values of a lock-nut. The two factors effecting the torque are the type of plating, and whether the locknut is threaded into a bolt or a mandrel. We'll use two-way ANOVA to find if the two factors or their interaction effect the torque.

```{r}
data <- read.table('bolt.dat.txt', header = TRUE)
data_melt <- melt(data, variable.name = 'plating', value.name = 'torque')
head(data_melt)
anova(lm(torque ~ M.B*plating, data = data_melt))
#data['plating_bolt'] = apply(data_melt[,1:2], axis = 1, function(x) paste)
```

The two factors and their interaction significantly effects the torque.

## Two-way layout: Random effects

```{r}
data <- read.csv('sensory_data.csv', header = TRUE)
head(data)
data_melt <- melt(data, variable.name = "Operator", value.name = 'property')
head(data_melt)
anova(lm(property~Item*Operator, data = data_melt))
```

Let us estimate the item variance:

```{r}
(68.04 - 0.36)/15
```

Let us estimate the operator variance:

```{r}
(6.37 - 0.36)/30
```
