# Multiple factors

## Randomized complete block design (RCBD)

Let us consider an example, where the shear strength of steel plate girders needs to be modeled as a function of the four methods *(treatment)* and nine girders *(blocks)*.

```{r}
library(ACSWR)
library(reshape2)
library(DescTools)

# Visualizing treatment effects
boxplot(girder[,2:5])

# Melting data to make it suitable to fit a linear regression model
# using the 'lm' function
data_melt <- melt(girder, variable.name = "treatment", id = 'Girder')
anova(lm(value~Girder+block, data = data_melt))
```

The null hypothesis that there is no effect of the method on the sheer strength of the girder is rejected.

Let us compare the methods pairwise.

```{r}
aov_object <- aov(value~Girder+block, data = data_melt)

# Tukey's method
PostHocTest(aov_object, method = "hsd")
```

All methods are different from each other according to Tukey's method *(at 5\% significance level)*.


```{r}
# Bonferroni's method
PostHocTest(aov_object, method = "bonferroni")
```

All methods except Aarau and Cardiff are different from each other according to Bonferroni's method *(at 5\% significance level)*.
