# Multiple factors

## Randomized complete block design (RCBD)

Let us consider an example, where the shear strength of steel plate girders needs to be modeled as a function of the four methods *(treatment)* and nine girders *(blocks)*.

```{r, message=FALSE}
library(ACSWR)
library(reshape2)
library(DescTools)
library(lmtest)

# Visualizing treatment effects
data("girder")
boxplot(girder[,2:5])

# Melting data to make it suitable to fit a linear regression model
# using the 'lm' function
data_melt <- melt(girder, variable.name = "treatment", id = 'Girder')
anova(lm(value~Girder+treatment, data = data_melt))
```

The null hypothesis that there is no effect of the method on the sheer strength of the girder is rejected.

Let us compare the methods pairwise.

```{r}
aov_object <- aov(value~Girder+treatment, data = data_melt)

# Tukey's method
PostHocTest(aov_object, method = "hsd")$treatment
```

All methods are different from each other according to Tukey's method *(at 5\% significance level)*.


```{r}
# Bonferroni's method
PostHocTest(aov_object, method = "bonferroni")$treatment
```

All methods except Aarau and Cardiff are different from each other according to Bonferroni's method *(at 5\% significance level)*. 

Let us verify if the model assumptions are satisfied.

```{r}
par(mfrow = c(2,2))
model <- lm(value~Girder+treatment, data = data_melt)
plot(model)
```


```{r}
shapiro.test(model$residuals)
```

The errors are normally distribued.

```{r}
bptest(model)
```

The error variance assumption is also satisfied at a 1\% significance level. There is no strong deviation from the assumption.


## Two-way layout: Fixed effects

A manufacturer was interested in finding differences in torque values of a lock-nut. The two factors effecting the torque are the type of plating, and whether the locknut is threaded into a bolt or a mandrel. We'll use two-way ANOVA to find if the two factors or their interaction effect the torque.

```{r}
data <- read.table('bolt.dat.txt', header = TRUE)
data_melt <- melt(data, variable.name = 'plating', value.name = 'torque')
head(data_melt)
anova(lm(torque ~ M.B*plating, data = data_melt))
#data['plating_bolt'] = apply(data_melt[,1:2], axis = 1, function(x) paste)
```

The two factors and their interaction significantly effects the torque.

## Two-way layout: Random effects

Ten food items are being tested by 5 judges *(operators)* for quality. Each judge inspects a food item 3 times, and gives a score. Answer the following questions with appropriate analysis:

1. Is there a statistically significant variation in the scores given by judges for the same quality of food? If yes, quantify the variation.

2. Is the variation in the scores given by judges for the same quality of food dependent on the food item?

3. Is there a statistically significant variation in the quality of the food items after removing the variation in scores due to different judges? If yes, quantify the variation.

```{r}
data <- read.csv('sensory_data.csv', header = TRUE)
head(data)
data_melt <- melt(data, variable.name = "Operator", value.name = 'property')
head(data_melt)
anova(lm(property~Item*Operator, data = data_melt))
```

Yes, there is a statistically significant variation in the scores given by judges for the same quality of food

The standard deviation in the scores given by judges for the same quality of food is:

```{r}
sqrt((6.37 - 0.36)/30)
```

The variation in the scores given by judges for the same quality of food does not depend on the food item

Yes, there is a statistically significant variation in the quality of the food items after removing the variation in scores due to different judges.

The standard deviation in the scores given food items after removing the variation due to different judges is:

```{r}
sqrt((68.04 - 0.36)/15)
```

