# Variable interactions

In this chapter, we'll introduce interactions, and then develop and interpret models with interactions among quantitative predictors, and interactions between quantitative and qualitative predictors.

```{r}
#loading libraries
library(ggplot2)
```


## Interaction model

Let us consider the file *car_data.csv*. 

```{r}
# Reading data
car_data <- read.csv('./Datasets/car_data.csv')
head(car_data)
```

In an additive model, we assume that the association between a predictor $X_j$ and response $Y$ does not depend on the value of other predictors. For example, consider the multiple linear regression model below.

```{r}
# Additive model
additive_model <- lm(price~year+engineSize+mileage+mpg, data = car_data)
summary(additive_model)
```

The above model assumes that the average increase in price associated with a unit increase in `engineSize` is always $12,180, regardless of the value of other predictors. However, this assumption may be incorrect.

## Interaction between continuous predictors

We can relax this assumption by considering another predictor, called an interaction term. Let us assume that the average increase in `price` associated with a one-unit increase in `engineSize` depends on the model `year` of the car. In other words, there is an interaction between `engineSize` and `year`. This interaction can be included as a predictor, which is the product of `engineSize` and `year`. *Note that there are several possible interactions that we can consider. Here the interaction between `engineSize` and `year` is just an example.*

```{r}
# Interaction model
interaction_model <- lm(price~year*engineSize+mileage+mpg, data = car_data)
summary(interaction_model)
```

Note that the $R^2$ has increased as compared to the additive model, since we added a predictor.

The model equation is:

`price` = $\beta_0$ + $\beta_1$`year` + $\beta_2$`engineSize` + $\beta_3$(`year` $\times$ `engineSize`) + $\beta_4$`mileage` + $\beta_5$`mpg`,
or

`price` = $\beta_0$ + $\beta_1$`year` + ($\beta_2+\beta_3$`year`) $\times$ `engineSize` + $\beta_4$`mileage` + $\beta_5$`mpg`,
or

`price` = $\beta_0 + \beta_1$`year` + $\tilde \beta$`engineSize` + $\beta_4$`mileage` + $\beta_5$`mpg`,


Since $\tilde \beta$ is a function of `year`, the association between `engineSize` and `price` is no longer a constant. A change in the value of `year` will change the association between `price` and `engineSize`.

Substituting the values of the coefficients:

`price` = 5.606e5 - 275.3833`year` + (-1.796e6+896.7687`year`)`engineSize` -0.1525`mileage` -84.3417`mpg`

Thus, for cars launched in the year 2010, the average increase in price for one liter increase in engine size is -1.796e6 + 896.7687 * 2010 $\approx$ \$6,500, assuming all the other predictors are constant. However, for cars launched in the year 2020, the average increase in price for one liter increase in engine size is -1.796e6 + 896.7687*2020 $\approx$ \$15,500 , assuming all the other predictors are constant.

Similarly, the equation can be re-arranged as:

`price` = 5.606e5 +(-275.3833+896.7687`engineSize`)`year` -1.796e6`engineSize` -0.1525`mileage` -84.3417`mpg`

Thus, for cars with an engine size of 2 litres, the average increase in price for a one year newer model is -275.3833+896.7687 * 2 $\approx$ \$1500, assuming all the other predictors are constant. However, for cars with an engine size of 3 litres, the average increase in price for a one year newer model is -275.3833+896.7687 * 3 $\approx$ \$2400, assuming all the other predictors are constant.

## Qualitative predictors

Let us develop a model for predicting `price` based on `engineSize` and the qualitative predictor `transmission`.

```{r}
#checking the distribution of values of transmission
table(car_data$transmission)
```

Note that the `Other` category of the variable `transmission` contains only a single observation, which is likely to be insufficient to train the model. We'll remove that observation from the car data. Another option may be to combine the observation in the `Other` category with the nearest category, and keep it in the data.

```{r}
car_data = car_data[car_data$transmission!='Other',]
```

```{r}
qual_pred_model <- lm(price~engineSize+transmission, data = car_data)
summary(qual_pred_model)
```

Note that there is no coefficient for the `Automatic` level of the variable `Transmission`. If a car doesn't have `Manual` or `Semi-Automatic` transmission, then it has an `Automatic` transmission. Thus, the coefficient of `Automatic` will be redundant, and the dummy variable corresponding to `Automatic` transmission is dropped from the model. 

The level of the categorical variable that is dropped from the model is called the baseline level. Here `Automatic` transmission is the baseline level. The coefficients of other levels of `transmission` should be interpreted with respect to the baseline level. 

**Q:** Interpret the intercept term.

**Ans:** For the hypothetical scenario of a car with zero engine size and `Automatic` transmission, the estimated mean car price is $\approx$ \$3042.

**Q:** Interpret the coefficient of `transmissionManual`.

**Ans:** The estimated mean price of a car with manual transmission is $\approx$ \$6770 less than that of a car with *Automatic* transmission.

Let us visualize the developed model.

```{r}
colors <- c("Automatic" = "red", "Manual" = "blue", "Semi-Automatic" = "green")

coefs <- qual_pred_model$coefficients
x <- car_data$engineSize
ggplot(data = car_data, aes(x = engineSize))+
  geom_line(aes(y = coefs['(Intercept)']+x*coefs['engineSize'], color = 'Automatic'))+
  geom_line(aes(y = coefs['(Intercept)']+x*coefs['engineSize']+coefs['transmissionManual'], color = 'Manual'))+
  geom_line(aes(y = coefs['(Intercept)']+x*coefs['engineSize']+coefs['transmissionSemi-Auto'], color = 'Semi-Automatic'))+
  theme(legend.title = element_blank(),
        legend.position = c(0.15,0.85))+
  labs(
    y = 'Predicted car price',
    x = 'Engine size (in litre)'
  )
```

Based on the developed model, for a given engine size, the car with a semi-automatic transmission is estimated to be the most expensive on average, while the car with a manual transmission is estimated to be the least expensive on average.

## Interaction between qualitative and continuous predictors

Note that the qualitative predictor leads to fitting 3 parallel lines to the data, as there are 3 categories.

However, note that we have made the constant association assumption. The fact that the lines are parallel means that the average increase in car price for one litre increase in engine size does not depend on the type of transmission. This represents a potentially serious limitation of the model, since in fact a change in engine size may have a very different association on the price of an automatic car versus a semi-automatic or manual car.

This limitation can be addressed by adding an interaction variable, which is the product of `engineSize` and the dummy variables for semi-automatic and manual transmissions. 

```{r}
qual_pred_int_model <- lm(price~engineSize*transmission, data = car_data)
summary(qual_pred_int_model)
```

The model equation for the model with interactions is:

Automatic transmission: `price` = 3754.7238	 + 9928.6082`engineSize`, 

Semi-Automatic transmission: `price` = 3754.7238 + 9928.6082`engineSize` + (-5282.7164+4162.2428`engineSize`), 

Manual transmission: `price` = 3754.7238 + 9928.6082`engineSize` +(1768.5856-5285.9059`engineSize`), or

Automatic transmission: `price` = 3754.7238	+ 9928.6082`engineSize`, 

Semi-Automatic transmission: `price` = -1527 + 7046`engineSize`, 

Manual transmission: `price` = 5523 + 4642`engineSize`.

**Q:** Interpret the coefficient of manual tranmission, i.e., the coefficient of `transmissionManual`.

**A:** For a given engine size, the estimated mean `price` of a car with `Manual` transmission is $\approx$ \$1768 more than the estimated mean `price` of a car with `Automatic` transmission.

**Q:** Interpret the coefficient of the interaction between engine size and manual transmission, i.e., the coefficient of `engineSize:transmissionManual`.

**A:** For a unit (or a litre) increase in `engineSize` , the increase in estimated mean `price` of a car with `Manual` transmission is $\approx$ \$5285 less than the increase in estimated mean `price` of a car with `Automatic` transmission.


```{r}
colors <- c("Automatic" = "red", "Manual" = "blue", "Semi-Automatic" = "green")

coefs <- qual_pred_int_model$coefficients
x <- car_data$engineSize
ggplot(data = car_data, aes(x = engineSize))+
  geom_line(aes(y = coefs['(Intercept)']+x*coefs['engineSize'], color = 'Automatic'))+
  geom_line(aes(y = coefs['(Intercept)']+x*coefs['engineSize']+coefs['transmissionManual']+x*coefs['engineSize:transmissionManual'], color = 'Manual'))+
  geom_line(aes(y = coefs['(Intercept)']+x*coefs['engineSize']+coefs['transmissionSemi-Auto']+x*coefs['engineSize:transmissionSemi-Auto'], color = 'Semi-Automatic'))+
  theme(legend.title = element_blank(),
        legend.position = c(0.15,0.85))+
  labs(
    y = 'Predicted car price',
    x = 'Engine size (in litre)'
  )
```


Note the interaction term adds flexibility to the model.

The slope of the regression line for semi-automatic cars is the largest. This suggests that increase in engine size is associated with a higher increase in car price for semi-automatic cars, as compared to other cars.