# Variable interactions

In this chapter, we'll introduce interactions, and then develop and interpret models with interactions among quantitative predictors, and interactions between quantitative and qualitative predictors.

## Interaction model

Let us consider the file *car_data.csv*. 

```{r}
# Reading data
car_data <- read.csv('./Datasets/car_data.csv')
head(car_data)
```

In an additive model, we assume that the association between a predictor $X_j$ and response $Y$ does not depend on the value of other predictors. For example, consider the multiple linear regression model below.

```{r}
# Additive model
additive_model <- lm(price~year+engineSize+mileage+mpg, data = car_data)
summary(additive_model)
```

The above model assumes that the average increase in price associated with a unit increase in `engineSize` is always $12,180, regardless of the value of other predictors. However, this assumption may be incorrect.

## Variable interaction between continuous predictors

We can relax this assumption by considering another predictor, called an interaction term. Let us assume that the average increase in `price` associated with a one-unit increase in `engineSize` depends on the model `year` of the car. In other words, there is an interaction between `engineSize` and `year`. This interaction can be included as a predictor, which is the product of `engineSize` and `year`. *Note that there are several possible interactions that we can consider. Here the interaction between `engineSize` and `year` is just an example.*

```{r}
# Interaction model
interaction_model <- lm(price~year*engineSize+mileage+mpg, data = car_data)
summary(interaction_model)
```

Note that the $R^2$ has increased as compared to the additive model, since we added a predictor.

The model equation is:

`price` = $\beta_0$ + $\beta_1$`year` + $\beta_2$`engineSize` + $\beta_3$(`year` $\times$ `engineSize`) + $\beta_4$`mileage` + $\beta_5$`mpg`,
or

`price` = $\beta_0$ + $\beta_1$`year` + ($\beta_2+\beta_3$`year`) $\times$ `engineSize` + $\beta_4$`mileage` + $\beta_5$`mpg`,
or

`price` = $\beta_0 + \beta_1$`year` + $\tilde \beta$`engineSize` + $\beta_4$`mileage` + $\beta_5$`mpg`,


Since $\tilde \beta$ is a function of `year`, the association between `engineSize` and `price` is no longer a constant. A change in the value of `year` will change the association between `price` and `engineSize`.

Substituting the values of the coefficients:

`price` = 5.606e5 - 275.3833`year` + (-1.796e6+896.7687`year`)`engineSize` -0.1525`mileage` -84.3417`mpg`

Thus, for cars launched in the year 2010, the average increase in price for one liter increase in engine size is -1.796e6 + 896.7687 * 2010 $\approx$ \$6,500, assuming all the other predictors are constant. However, for cars launched in the year 2020, the average increase in price for one liter increase in engine size is -1.796e6 + 896.7687*2020 $\approx$ \$15,500 , assuming all the other predictors are constant.

Similarly, the equation can be re-arranged as:

`price` = 5.606e5 +(-275.3833+896.7687`engineSize`)`year` -1.796e6`engineSize` -0.1525`mileage` -84.3417`mpg`

Thus, for cars with an engine size of 2 litres, the average increase in price for a one year newer model is -275.3833+896.7687 * 2 $\approx$ \$1500, assuming all the other predictors are constant. However, for cars with an engine size of 3 litres, the average increase in price for a one year newer model is -275.3833+896.7687 * 3 $\approx$ \$2400, assuming all the other predictors are constant.

## Qualitative predictors

Let us develop a model for predicting `price` based on `engineSize` and the qualitative predictor `transmission`.

```{r}
#checking the distribution of values of transmission
table(car_data$transmission)
```

Note that the `Other` category of the variable `transmission` contains only a single observation, which is likely to be insufficient to train the model. We'll remove that observation from the car data. Another option may be to combine the observation in the `Other` category with the nearest category, and keep it in the data.

```{r}
car_data = car_data[car_data$transmission!='Other',]
```

```{r}
qual_pred_model <- lm(price~engineSize+transmission, data = car_data)
summary(qual_pred_model)
```

Note that there is no coefficient for the `Automatic` level of the variable `Transmission`. If a car doesn't have `Manual` or `Semi-Automatic` transmission, then it has an `Automatic` transmission. Thus, the coefficient of `Automatic` will be redundant, and the dummy variable corresponding to `Automatic` transmission is dropped from the model. 

The level of the categorical variable that is dropped from the model is called the baseline level. Here `Automatic` transmission is the baseline level. The coefficients of other levels of `transmission` should be interpreted with respect to the baseline level. 

**Q:** Interpret the intercept term.

**Ans:** For the hypothetical scenario of a car with zero engine size and `Automatic` transmission, the estimated mean car price is $\approx$ \$3042.

**Q:** Interpret the coefficient of `transmissionManual`.

**Ans:** The estimated mean price of a car with manual transmission is $\approx$ \$6770 less than that of a car with *Automatic* transmission.

Let us visualize the developed model.




