# Quantitative & Qualitative predictors

In this chapter, we'll introduce interactions, and then develop and interpret models with interactions among quantitative predictors, and interactions between quantitative and qualitative predictors.

## Variable interactions

Let us consider the file *car_data.csv*. 

```{r}
# Reading data
car_data <- read.csv('./Datasets/car_data.csv')
head(car_data)
```

In an additive model, we assume that the association between a predictor $X_j$ and response $Y$ does not depend on the value of other predictors. For example, consider the multiple linear regression model below.

```{r}
# Additive model
additive_model <- lm(price~year+engineSize+mileage+mpg, data = car_data)
summary(additive_model)
```

The above model assumes that the average increase in price associated with a unit increase in `engineSize` is always $12,180, regardless of the value of other predictors. However, this assumption may be incorrect.

### Variable interaction between continuous predictors

We can relax this assumption by considering another predictor, called an interaction term. Let us assume that the average increase in `price` associated with a one-unit increase in `engineSize` depends on the model `year` of the car. In other words, there is an interaction between `engineSize` and `year`. This interaction can be included as a predictor, which is the product of `engineSize` and `year`. *Note that there are several possible interactions that we can consider. Here the interaction between `engineSize` and `year` is just an example.*

```{r}
# Interaction model
interaction_model <- lm(price~year*engineSize+mileage+mpg, data = car_data)
summary(interaction_model)
```

Note that the $R^2$ has increased as compared to the additive model, since we added a predictor.

The model equation is:

`price` = $\beta_0$ + $\beta_1$`year` + $\beta_2$`engineSize` + $\beta_3$(`year` $\times$ `engineSize`) + $\beta_4$`mileage` + $\beta_5$`mpg`,
or

`price` = $\beta_0$ + $\beta_1$`year` + ($\beta_2+\beta_3$)`year` $\times$ `engineSize` + $\beta_4$`mileage` + $\beta_5$`mpg`,
or

`price` = $\beta_0 + \beta_1$`year` + $\tilde \beta$`engineSize` + $\beta_4$`mileage` + $\beta_5$`mpg`,


Since $\tilde \beta$ is a function of `year`, the association between `engineSize` and `price` is no longer a constant. A change in the value of `year` will change the association between `price` and `engineSize`.

Substituting the values of the coefficients:

`price` = 5.606e5 - 275.3833`year` + (-1.796e6+896.7687`year`)`engineSize` -0.1525`mileage` -84.3417`mpg`





