# Hypothesis testing example

## Question

Cola manufacturers want to test how much the sweetness of a new cola drink is affected by storage. The sweetness loss due to storage was evaluated by 10 professional tasters (by comparing the sweetness before and after storage):

```         
Taster          Sweetness loss

 1         2.0
 2         0.4
 3         0.7  
 4         2.0  
 5       −0.4   
 6         2.2  
 7       −1.3   
 8         1.2  
 9         1.1
10         2.3
```

Obviously, we want to test if storage results in a loss of sweetness

Let $\mu$ denote the sweetness loss, thus:

Null hypotheis: H_0: \mu = 0

Alternate hypotheis: $H_a: \mu > 0$

## Solution

Sample mean ($\bar{x}$):

```{r}
data <- c(2, 0.4, 0.7, 2, -0.4, 2.2, -1.3, 1.2, 1.1, 2.3)

xbar <- mean(data)
xbar
```

T-statistic:

```{r}
t = xbar/(sd(data)/sqrt(10))
t
```

p-value:

```{r}
1-pt(t, df = 9)
```

If the probability of Type I error considered is 5%, then we reject the null hypothesis, and conclude that the sweetness loss is indeed greater than 0.

If the probability of Type I error considered is 1%, then we fail to reject the null hypothesis, and conclude that the sweetness loss is indeed 0.
