# Hypothesis testing examples

## Question

Cola manufacturers want to test how much the sweetness of a new cola drink is affected by storage. The sweetness loss due to storage was evaluated by 10 professional tasters (by comparing the sweetness before and after storage):

```         
Taster          Sweetness loss

 1         2.0
 2         0.4
 3         0.7  
 4         2.0  
 5       −0.4   
 6         2.2  
 7       −1.3   
 8         1.2  
 9         1.1
10         2.3
```

Obviously, we want to test if storage results in a loss of sweetness

Let $\mu$ denote the sweetness loss, thus:

Null hypothesis: $H_0: \mu = 0$

Alternate hypothesis: $H_a: \mu > 0$

## Solution

Sample mean ($\bar{x}$):

```{r}
data <- c(2, 0.4, 0.7, 2, -0.4, 2.2, -1.3, 1.2, 1.1, 2.3)

xbar <- mean(data)
xbar
```

T-statistic:

```{r}
t = xbar/(sd(data)/sqrt(10))
t
```

p-value:

```{r}
1-pt(t, df = 9)
```

If the probability of Type I error considered is 5%, then we reject the null hypothesis, and conclude that the sweetness loss is indeed greater than 0.

If the probability of Type I error considered is 1%, then we fail to reject the null hypothesis, and conclude that the sweetness loss is indeed 0.

## Question

In a study of lettuce growth, ten seedlings were randomly allocated to be grown in either a standard nutrient solution or in a solution containing extra nitrogen. After 22 days, the plants were harvested and weighed. The table below summarizes the results. Can we conclude that extra nitrogen enhances growth?

| Nutrient solution | n   | mean |  SD  |
|-------------------|:----|-----:|:----:|
| Standard          | 5   | 3.62 | 0.54 |
| Extra             | 5   | 4.17 | 0.67 |

## Solution

We will first test the hypothesis if the variance of the responses corresponding to the two treatments are the same or not.

We will assume that the response follows the normal distribution.

Let $\mu$ denote the sweetness loss, thus:

Null hypothesis: $H_0: \mu = 0$

Alternate hypothesis: $H_a: \mu > 0$



